---
title: "LM_Model"
author: "Amar Wadhwani"
date: "2025-06-08"
output: html_document
---
```{r}

ensureLibrary <- function(pkg) {
  suppressPackageStartupMessages({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg)
    }
    library(pkg, character.only = TRUE)
  })
}

ensureLibrary("biglm")
ensureLibrary("slurmR")
ensureLibrary("doParallel")
ensureLibrary("parallel")
ensureLibrary("caret")
ensureLibrary("e1071")
ensureLibrary("ROCR")
ensureLibrary("httr")
ensureLibrary("jsonlite")
ensureLibrary("dplyr")
ensureLibrary("tidyr")
ensureLibrary("lubridate")
ensureLibrary("ggplot2")
ensureLibrary("readr")
ensureLibrary("caret")
ensureLibrary("purrr")
ensureLibrary("ggplot2")
ensureLibrary("neuralnet")
ensureLibrary("xgboost")
ensureLibrary("randomForest")
ensureLibrary("patchwork")
ensureLibrary("Metrics")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lm_model}
df <- read_csv("./eda/taxi_fare_model_data.csv")
str(df)
nrow(df)
```

We have 2 million rows for the year 2024, which have pre-processed during our EDA.
We will now select the relevant features, convert them to factors where appropriate.
```{r}
df_model <- df %>%
  select(
    fare,
    trip_hour,
    trip_day,
    trip_month,
    is_weekend,
    is_rush_hour,
    trip_duration,
    trip_miles,
    pickup_community_area,
    dropoff_community_area,
    pickup_census_tract,
    dropoff_census_tract
  ) %>%
  mutate(
    trip_day = as.factor(trip_day),
    trip_month = as.factor(trip_month),
    is_weekend = as.factor(is_weekend),
    is_rush_hour = as.factor(is_rush_hour),
    pickup_community_area = as.factor(pickup_community_area),
    dropoff_community_area = as.factor(dropoff_community_area),
    pickup_census_tract = as.factor(pickup_census_tract),
    dropoff_census_tract = as.factor(dropoff_census_tract)
  ) %>%
  na.omit()
```

We now train the model on train dataset and will test the predictions
```{r}
set.seed(42)
factor_vars <- c(
  "trip_day", "trip_month", "is_weekend", "is_rush_hour",
  "pickup_community_area", "dropoff_community_area",
  "pickup_census_tract", "dropoff_census_tract"
)
df_sample <- df_model %>% sample_frac(0.5)
getTrainAndTestSamples_BalancedFactors <- function(data, percentTrain = 0.75, seed = 42, replaceIfRare = TRUE) {
  set.seed(seed)

  if (!is.data.frame(data)) stop("Input `data` must be a data frame.")

  factor_vars <- names(Filter(is.factor, data))
  
  # Initialize train and test sets
  train_list <- list()
  test_list <- list()

  # For each factor column, ensure all levels appear in both sets
  for (col in factor_vars) {
    levels_col <- levels(data[[col]])
    
    for (lvl in levels_col) {
      # Subset rows where this level occurs
      rows_with_level <- data[data[[col]] == lvl, ]
      n <- nrow(rows_with_level)

      if (n == 1) {
        # Only one occurrence — put it in both sets if allowed
        if (replaceIfRare) {
          train_list[[length(train_list) + 1]] <- rows_with_level
          test_list[[length(test_list) + 1]] <- rows_with_level
        } else {
          warning(sprintf("Skipping rare level '%s' in column '%s' (only 1 instance)", lvl, col))
        }
      } else {
        # Sample index for train set
        idx <- sample(seq_len(n), size = floor(n * percentTrain), replace = FALSE)
        train_list[[length(train_list) + 1]] <- rows_with_level[idx, ]
        test_list[[length(test_list) + 1]] <- rows_with_level[-idx, ]
      }
    }
  }

  # Combine factor-balanced subsets
  train_df <- do.call(rbind, train_list)
  test_df  <- do.call(rbind, test_list)

  # Remove potential duplicate rows introduced by overlapping rare levels
  train_df <- unique(train_df)
  test_df  <- unique(test_df)

  # Add back any remaining rows not included above
  all_ids <- paste0(rownames(train_df), "_train") # to track uniqueness
  all_ids <- c(all_ids, paste0(rownames(test_df), "_test"))

  rest <- data[!rownames(data) %in% unique(c(rownames(train_df), rownames(test_df))), ]
  rest_n <- nrow(rest)
  if (rest_n > 0) {
    idx_rest <- sample(seq_len(rest_n), size = floor(rest_n * percentTrain))
    train_df <- rbind(train_df, rest[idx_rest, ])
    test_df  <- rbind(test_df, rest[-idx_rest, ])
  }

  return(list(train = train_df, test = test_df))
}

stratified_train_test_split <- function(data, target_frac = 0.8,
                                        factor_cols = NULL,
                                        numeric_cols = NULL,
                                        bins = 4,
                                        min_stratum_size = 5,
                                        seed = 42) {
  set.seed(seed)

  df <- data

  # Combine factor and binned numeric columns to create strata
  if (!is.null(factor_cols)) {
    for (col in factor_cols) {
      df[[col]] <- as.factor(df[[col]])
    }
  }

  if (!is.null(numeric_cols)) {
    for (col in numeric_cols) {
      # Bin numeric columns into quantile bins
      binned <- cut(df[[col]], breaks = quantile(df[[col]], probs = seq(0, 1, length.out = bins + 1),
                                                 na.rm = TRUE), include.lowest = TRUE, labels = FALSE)
      df[[paste0(col, "_bin")]] <- as.factor(binned)
    }
  }

  # Create stratification key by combining all columns
  strat_cols <- c(factor_cols, paste0(numeric_cols, "_bin"))
  df$strata <- apply(df[strat_cols], 1, paste, collapse = "_")

  # Remove strata that are too small
  stratum_counts <- table(df$strata)
  valid_strata <- names(stratum_counts[stratum_counts >= min_stratum_size])
  df <- df[df$strata %in% valid_strata, ]

  # Stratified sampling
  train_indices <- unlist(lapply(split(1:nrow(df), df$strata), function(idxs) {
    n_train <- floor(length(idxs) * target_frac)
    sample(idxs, n_train)
  }))

  train <- df[train_indices, !(names(df) %in% c("strata", paste0(numeric_cols, "_bin")))]
  test  <- df[-train_indices, !(names(df) %in% c("strata", paste0(numeric_cols, "_bin")))]

  list(train = train, test = test)
}
# Use the function
#splits <- split_balanced(df_sample, factor_vars)
#train_data <- splits$train
#test_data <- splits$test
splits_2 <- stratified_train_test_split(
  data = df_sample,
  target_frac = 0.8,
  factor_cols = factor_vars,
  numeric_cols = c("trip_duration", "trip_miles", "trip_hour"),
  bins = 4, # number of bins for numeric columns
  min_stratum_size = 5 # minimum samples in a stratum to be kept
)

train_data_2 <- splits_2$train
test_data_2 <- splits_2$test

nrow(df_sample)
```
```{r}
# Sanity check for same levels in test-train split

for (var in factor_vars) {
  train_levels <- levels(train_data_2[[var]])
  test_levels <- levels(test_data_2[[var]])

  missing_in_train <- setdiff(levels(df_model[[var]]), train_levels)
  missing_in_test <- setdiff(levels(df_model[[var]]), test_levels)

  if (length(missing_in_train) > 0 || length(missing_in_test) > 0) {
    cat(paste0("Missing levels for ", var, "\n"))
    print(list(train_missing = missing_in_train, test_missing = missing_in_test))
  }
}
```
```{R eval=FALSE}
#cl <- makeSlurmCluster(6)
#registerDoParallel(cl)

#library(doParallel)
#registerDoParallel(cl)
```
```{r}
predictors <- setdiff(names(train_data_2), "fare")
form <- as.formula(paste("fare ~", paste(predictors, collapse = " + ")))

lm <- lm(form, data = train_data_2)

```

```{r}
# Predict on test data
predictions <- predict(lm, newdata = test_data_2)

# Evaluate performance
actuals <- test_data_2$fare
rmse <- sqrt(mean((predictions - actuals)^2))
mae <- mean(abs(predictions - actuals))
r2 <- 1 - sum((predictions - actuals)^2) / sum((actuals - mean(actuals))^2)

cat("RMSE:", rmse, "\nMAE:", mae, "\nR-squared:", r2, "\n")
```
For sample size 0.005% of cleaned data, the following metrics are observed
RMSE: 3.904028 
MAE: 1.728776 
R-squared: 0.9532892 

For sample size 0.05% of cleaned data, the following metrics are observed
RMSE: 4.16958 
MAE: 1.834244 
R-squared: 0.94698 
For sample size 0.5% of cleaned data without replacement, the following metrics are observed
RMSE: 3.580551 
MAE: 1.409825 
R-squared: 0.9655701
```{r}
df_plot <- data.frame(actual = actuals, predicted = predictions)

# Plot actual vs. predicted
ggplot(df_plot, aes(x = actual, y = predicted)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Predicted Fare",
       x = "Actual Fare",
       y = "Predicted Fare") +
  theme_minimal()
```
```{r eval = FALSE}
#stopCluster(cl)
```
